Adaptive evolution with aneuploidy and mutation

### 2021-06-22 ms/ms.tex
continuted with methods

\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
      \caption{
        \textbf{A. Single-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
      \label{fig:model1}
  \end{subfigure}%
  \\
  \begin{subfigure}[t]{0.5\textwidth}
  	  \caption{
        \textbf{B. Multi-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
      \label{fig:model2}
  \end{subfigure}
  \caption{
    \textbf{Model illustrations.}
    \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt to \eumt.
    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, mutation rate; $\alpha$, aneuploidy gain rate; $\alpha_l$, aneuploidy loss rate.
  }
  \label{fig:models}
\end{figure}

@@

%%% Table: single-locus model
\begin{table}[h]
\centering
\caption{\textbf{Single-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+\sigma$ & $(1-c)(1+s)+\sigma$ & $1+s$               
\end{tabular}
\label{table:single-locus}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$\sigma \ge 0$ is the selection coefficient of aneuploidy; 
and $0 \le c \le 1$ is the fitness cost of aneuploidy.
}
\end{table}
%%%

@@

A genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$, $s_i \sim Exp(s)$.
The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $\sigma$, $c$) are the same as in the single-locus model.
The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough. The intuition is that when the benefit of the accumulated beneficial mutations is small, the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations benefit is large, then aneuploidy is no longer advantageous because of its significant cost.

@@

%%% Table: multi-locus model
\begin{table}[h]
\centering
\caption{\textbf{Multi-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+\sigma$ & $(1-c)\prod_{j=1}^k(1+s_i)+\sigma$ & $\prod_{j=1}^k(1+s_i)$               
\end{tabular}
\label{table:multi-locus}
\caption*{
$k$ is the number of accumulated beneficial mutations in the genome;
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$\sigma \ge 0$ is the selection coefficient of aneuploidy; 
and $0 \le c \le 1$ is the fitness cost of aneuploidy.
}
\end{table}
%%%

\paragraph{Empirical evidence.}
Our inference procedure uses empirical data from evolutionary experiments performed by \citet{Yona2012}.
In the heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}. Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype means that 95\% or 5\% of the population carry the genotype).
The experiment continued with two populations, in which aneuploidy was eliminated by generation 1700 and 2350.
In the pH-stress experiment, four populations of \yeast evolved under high-ph stress (8.6). Aneuploidy fixed during the first 150 generations. It was fully eliminated in two populations and partly eliminated in two populations by generation 750.

\paragraph{Likelihood function.}
Denote by $A_{t}$ the fixation of aneuploidy at generation $t$, by $L_{t}$ the loss of aneuploidy at generation $t$, and by $L^*_{t,T}$ the loss of aneuploidy generation $t$ conditioned on no loss by generation $T$.
The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
\begin{multline} \label{eq:heatstress-likelihood}
\likelihood(\theta | A_{450}, L_{1700}, L^*_{2350, 1700} ) = \\
P^4(A_{450}) \cdot \Big(1 - P^4(\lnot{L_{1700}} | A_{450} ) - P^4(\lnot{L^*_{2350, 1700}}| A_{450}) + P^4(\lnot{L_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450})\Big)	\;,	
\end{multline}
where $P(X)$ is the probability of even $X$ and $\lnot$ means \emph{not}.
The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
\begin{equation} \label{eq:phstress-likelihood}
\likelihood(\theta | A_{150}, L_{750} ) = \\
 P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150}) \;.
\end{equation}

\paragraph{Parameter inference.} To estimate the parameters of the models we apply approximate Bayesian computation with a sequential Monte-Carlo scheme (ABC-SMC), employing pyabc Python 
package \citep{Klinger2018}

>>>>

\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
      \caption{
        \textbf{A. Single-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
      \label{fig:model1}
  \end{subfigure}%
  \\
  \begin{subfigure}[t]{0.5\textwidth}
  	  \caption{
        \textbf{B. Multi-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
      \label{fig:model2}
  \end{subfigure}
  \caption{
    \textbf{Model illustrations.}
    \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt to \eumt.
    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, mutation rate; $\delta$, aneuploidy gain rate; $\delta_L$, aneuploidy loss rate.
  }
  \label{fig:models}
\end{figure}

@@

%%% Table: single-locus model
\begin{table}[h]
\centering
\caption{\textbf{Single-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:single-locus}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$b \ge 0$ is the selection coefficient of aneuploidy; 
and $0 \le c \le 1$ is the fitness cost of aneuploidy.
}
\end{table}
%%%

@@

A genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$, $s_i \sim Exp(s)$.
The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $b$, $c$) are the same as in the single-locus model.
The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough. The intuition is that when the benefit of the accumulated beneficial mutations is small, the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations benefit is large, then aneuploidy is no longer advantageous because of its significant cost.

@@

%%% Table: multi-locus model
\begin{table}[h]
\centering
\caption{\textbf{Multi-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)\prod_{j=1}^k(1+s_i)+b$ & $\prod_{j=1}^k(1+s_i)$               
\end{tabular}
\label{table:multi-locus}
\caption*{
$k$ is the number of accumulated beneficial mutations in the genome;
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$b \ge 0$ is the selection coefficient of aneuploidy; 
and $0 \le c \le 1$ is the fitness cost of aneuploidy.
}
\end{table}
%%%

\paragraph{Empirical evidence.}
Our inference procedure uses empirical data from evolutionary experiments performed by \citet{Yona2012}.
In the heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}. Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype means that 95\% or 5\% of the population carry the genotype).
The experiment continued with two populations, in which aneuploidy was eliminated by generation 1700 and 2350.
In the pH-stress experiment, four populations of \yeast evolved under high-ph stress (8.6). Aneuploidy fixed during the first 150 generations. It was fully eliminated in two populations and partly eliminated in two populations by generation 750. 
These empirical results were published by \citet{Yona2012}.

\paragraph{Likelihood function.}
Denote by $A_{t}$ the fixation of aneuploidy at generation $t$, by $L_{t}$ the loss of aneuploidy at generation $t$, and by $L^*_{t,T}$ the loss of aneuploidy at generation $t$ conditioned on no loss by generation $T$.
The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
\begin{multline} \label{eq:heatstress-likelihood}
\likelihood(\theta | A_{450}, L_{1700}, L^*_{2350, 1700} ) = \\
P^4(A_{450}) \cdot \Big(1 - P^4(\lnot{L_{1700}} | A_{450} ) - P^4(\lnot{L^*_{2350, 1700}}| A_{450}) + P^4(\lnot{L_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450})\Big)	\;,	
\end{multline}
where $P(X)$ is the probability of even $X$ and $\lnot$ means \emph{not}.
The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
\begin{equation} \label{eq:phstress-likelihood}
\likelihood(\theta | A_{150}, L_{750} ) = \\
 P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150}) \;.
\end{equation} 
Because of the complexity of the model, this likelihood function is intractable. Therefore, given specific values of the model parameters $\theta$, we simulate the model in many replicated to approximate the value of the likelihood function $\likelihood(\theta)$.

\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyabc} Python 
package~\citep{Klinger2018}.
Briefly, this approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters. This approach is applicable for likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be computed. It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed. 
% TODO ILIA please add all configuration details, similar to the relevant section in the NPI paper. Discuss priors, convergence tests, etc.

\paragraph{Prior distributions.} The prior distributions of $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$ were obtained by estimating prior distributions for $1-c$, $1-c+b$ and $1+s$ from growth curves data. These growth curve data were previously obtained by~\citet{Yona2012} using mono-culture growth experiments.
The raw growth curves data were not published before, but were used to produce figures 3C, 4A, and S2 in \citet{Yona2012}.
We used \texttt{Curveball}, a dedicated method for predicting results of competition experiments from growth curve data~\citet{Ram2019}. \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed-culture, that is, it predicts the results of a competition assay. From these predictions, relative fitness values can be computed.  
We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} and in \SI{30}{\celsius} to estimate $1-c+b$ and $1-c$, respectively. The latter assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might not be correct, but we only assume this in order to generate a prior distribution for the fitness values. We also used growth curves of $2n+1$ and $2n^*$ in \SI{39}{\celsius} in \SI{39}{\celsius} to estimate $1+s$. 
Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. Thus, we sampled 10,000 values for $1-c+b$, $1-c$, and $1+s$, which we used as prior distributions. Note that the first and last values correspond to $w_{2n+1}$ and $w_{2n}$, whereas $w_{2n+1^*}=(1-c)w_{2n^*}+w_{2n+1}-(1-c)$.
%TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
%TODO ILIA describe the priors for the mutation and aneuploidy rate.

\paragraph{Model comparison.}  % TODO ILIA what did you do for model comparison/selection?
